Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-03-31 article/06_Supplementary_Remarks.Rmd
phrasing

4. ^14^C dating requires a small but sufficiently contextualised amount of organic finds. Due to the old wood and old water effect, short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this paper comes from bones and teeth. However, their preservation is subject to huge differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This situation can affect entire regions, if, for example, poor bone preservation prevails as a result of lime-deficient soils.

@@

The disturbance and robbery of graves was a widespread phenomenon in history and prehistory. Burial mounds were usually attacked more severely than flat graves due to their better visibility and could, therefore, be under-represented in Radon-B. In many places, they were also the subject of expeditions in the early days of archaeology at the end of the 19th and beginning of the 20th century. Of course, these excavations did not yield ^14^C-dates. Irrespective of this, it can be assumed that a large part of the burial mounds erected in the Bronze Age was destroyed by agricultural activity. Ploughing could also result in an under-representation of the archaeological findings [@hardingEuropeanSocietiesBronze2000a]. On the other hand, if a burial mound is still present, it is more readily visible and recognised by archaeologists today: Mounds are often clearly identifiable in aerial photographs or digital terrain models, while flat graves are much more difficult to identify for today's remote sensing methods. It is difficult to estimate the impact of these effects in detail, but an over-representation of mounds seems more likely than an over-representation of flat graves.

@@

Another aspect to consider concerning the data quality is the temporal accuracy of the radiocarbon dating. Ideally each radiocarbon date represents the exact point in time of the creation of the respective grave. Unfortunately several natural and cultural effects can cause a (partially systematic) offset between the radiocarbon date and the burial ceremony, including the *Marine reservoir effect*, the *Old wood effect*, the *Human ^14^C-residence ("adult") effect* or cultural practices like a secondary treatment of the body or the deposition of burial objects that are already ancient at the time of the ceremony. These effects can have strong influence on the accuracy of the dating information, but can only be explored for each grave individually. In the present paper they therefore had to be ignored.

@@

This paper is based on ^14^C dates which were prepared for the dating of graves. There is a fundamental break between that great, theoretical archaeology, which wants to develop comprehensive narratives and only pay quantitative attention to the individual case, and the on-site research or emergency excavation, which has to dedicate a measure of respect and sometimes a lot of time to each grave. In this paper, ^14^C data are converted into a proxy of calibrated density distributions within artificial working regions, which is completely detached from the individual context. ^14^C data on graves are used to attribute a feature to the absolute chronological context or to better describe the relative development within a site. The questions underlying the creation of a ^14^C date are therefore not equivalent to those imposed in this paper.

>>>>

4. ^14^C dating requires a small but sufficiently contextualised amount of organic finds and short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this paper comes from bones and teeth. However, their preservation is subject to huge differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This situation can affect entire regions, if, for example, poor bone preservation prevails as a result of lime-deficient soils.

@@

The disturbance and robbery of graves was a widespread phenomenon in history and prehistory. Burial mounds were usually attacked more severely than flat graves due to their better visibility and could, therefore, be under-represented in Radon-B. In many places, they were also the subject of expeditions in the early days of archaeology at the end of the 19th and beginning of the 20th century. Of course, these excavations did not yield ^14^C-dates. Irrespective of this, it can be assumed that a large part of the burial mounds erected in the Bronze Age was destroyed by agricultural activity. Ploughing could also result in an under-representation of the archaeological findings [@hardingEuropeanSocietiesBronze2000a]. On the other hand, if a burial mound is still present, it is also more readily visible and recognised by archaeologists today: Mounds are often clearly identifiable in aerial photographs or digital terrain models, while flat graves are much more difficult to observe with modern remote sensing methods. It is difficult to estimate the impact of these effects in detail, but an over-representation of mounds seems more likely than an over-representation of flat graves.

@@

Another aspect to consider concerning the data quality is the temporal accuracy of radiocarbon dating. Ideally each radiocarbon date represents the exact point in time of the creation of the respective grave. Unfortunately several natural and cultural effects can cause a (partially systematic) offset between the radiocarbon date and the burial ceremony, including the *Marine reservoir effect* [@Ascough2005], the *Old wood effect* [@OLSEN201330], the *Human ^14^C-residence ("adult") effect* [@Hedges2007] or cultural practices like a secondary treatment of the body or the deposition of burial objects that are already ancient at the time of the funeral. These effects can have strong influence on the accuracy of the dating information, but can only be explored for each grave individually. In the present paper they therefore had to be ignored.

@@

This paper is based on ^14^C dates which were prepared for the dating of graves. There is a semantic divisiveness between great, theoretical archaeology, which wants to develop comprehensive narratives and only pay quantitative attention to the individual case, and the on-site research or emergency excavation, which has to dedicate a measure of respect and sometimes a lot of time to each grave.

In this paper, ^14^C data are converted into a proxy of calibrated density distributions within artificial working regions. An endeavour which is completely detached from the individual context. In contrast, a single ^14^C date on a grave is usually used to attribute a feature to the absolute chronological context or to better describe the relative development within a site. The questions underlying the creation of a ^14^C date are therefore fundamentally different from those imposed here.
